
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Red Wine Quality Analysis by Marko Horvat
========================================================
Citation Request:
  This dataset is public available for research. The details are described in [Cortez et al., 2009]. 
  Please include this citation if you plan to use this database:

  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  Available at: [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
                [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
                [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib

1. Title: Wine Quality 

2. Sources
   Created by: Paulo Cortez (Univ. Minho), Antonio Cerdeira, Fernando Almeida, Telmo Matos and Jose Reis (CVRVV) @ 2009
   
3. Past Usage:

  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

  In the above reference, two datasets were created, using red and white wine samples.
  The inputs include objective tests (e.g. PH values) and the output is based on sensory data
  (median of at least 3 evaluations made by wine experts). Each expert graded the wine quality 
  between 0 (very bad) and 10 (very excellent). Several data mining methods were applied to model
  these datasets under a regression approach. The support vector machine model achieved the
  best results. Several metrics were computed: MAD, confusion matrix for a fixed error tolerance (T),
  etc. Also, we plot the relative importances of the input variables (as measured by a sensitivity
  analysis procedure).
 
4. Relevant Information:

   The two datasets are related to red and white variants of the Portuguese "Vinho Verde" wine.
   For more details, consult: http://www.vinhoverde.pt/en/ or the reference [Cortez et al., 2009].
   Due to privacy and logistic issues, only physicochemical (inputs) and sensory (the output) variables 
   are available (e.g. there is no data about grape types, wine brand, wine selling price, etc.).

   These datasets can be viewed as classification or regression tasks.
   The classes are ordered and not balanced (e.g. there are munch more normal wines than
   excellent or poor ones). Outlier detection algorithms could be used to detect the few excellent
   or poor wines. Also, we are not sure if all input variables are relevant. So
   it could be interesting to test feature selection methods. 

5. Number of Instances: red wine - 1599; white wine - 4898. 

6. Number of Attributes: 11 + output attribute
  
   Note: several of the attributes may be correlated, thus it makes sense to apply some sort of
   feature selection.

7. Attribute information:

   For more information, read [Cortez et al., 2009].

   Input variables (based on physicochemical tests):
   1 - fixed acidity (tartaric acid - g / dm^3)
   2 - volatile acidity (acetic acid - g / dm^3)
   3 - citric acid (g / dm^3)
   4 - residual sugar (g / dm^3)
   5 - chlorides (sodium chloride - g / dm^3
   6 - free sulfur dioxide (mg / dm^3)
   7 - total sulfur dioxide (mg / dm^3)
   8 - density (g / cm^3)
   9 - pH
   10 - sulphates (potassium sulphate - g / dm3)
   11 - alcohol (% by volume)
   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)

8. Missing Attribute Values: None

9. Description of attributes:

   1 - fixed acidity: most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

   2 - volatile acidity: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

   3 - citric acid: found in small quantities, citric acid can add 'freshness' and flavor to wines

   4 - residual sugar: the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

   5 - chlorides: the amount of salt in the wine

   6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

   7 - total sulfur dioxide: amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

   8 - density: the density of water is close to that of water depending on the percent alcohol and sugar content

   9 - pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

   10 - sulphates: a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant

   11 - alcohol: the percent alcohol content of the wine

   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(knitr)
library(ggplot2)
library(varhandle)
library(GGally)
library(gridExtra)
library(dplyr)
library(memisc)

```

```{r echo=FALSE}
wqr_raw <- read.csv('wineQualityReds.csv')
```


```{r echo=FALSE}
ncol(wqr_raw)
nrow(wqr_raw)
```


```{r echo=FALSE}
str(wqr_raw)
```


```{r echo=FALSE}
wqr <- within(wqr_raw, rm(X))
summary(wqr)
```


```{r echo=FALSE}
ggplot(aes(x = as.factor(quality)), data = wqr)+ geom_bar()
summary(wqr$quality)
```

From the bar plot above it can be observed that distribution for quality is almost normally distributed


```{r echo=FALSE}
ggplot(data = wqr, aes(x = fixed.acidity)) +
geom_histogram(binwidth = .1)

summary(wqr$fixed.acidity)
```

From the histogram above it can be observed that distribution for fixed acidity is right skewed wiht few outliers


```{r echo=FALSE}
ggplot(data = wqr, aes(x = volatile.acidity)) +
geom_histogram(binwidth = 0.015)

summary(wqr$volatile.acidity)
```

From the histogram above it can be observed that volatile acidity is normally distributed wiht few outliers


```{r echo=FALSE}
ggplot(data = wqr, aes(x = citric.acid)) +
geom_histogram(binwidth = 0.01)

summary(wqr$citric.acid)
```

Citric acid appears to have uniformal distribution wiht a large number of zero values.


```{r echo=FALSE}
ggplot(data = wqr, aes(x = residual.sugar)) +
geom_histogram(binwidth = 0.15)

summary(wqr$residual.sugar)
```

From the histogram above it can be observed that residual sugar is normally distributed with enormous number of outliners.


```{r echo=FALSE}
ggplot(data = wqr, aes(x = chlorides)) +
geom_histogram(binwidth = 0.005)

summary(wqr$chlorides)
```

From the histogram above it can be observed that chlorides is normally distributed with enormous number of outliners.


```{r echo=FALSE}
ggplot(data = wqr, aes(x = free.sulfur.dioxide)) +
geom_histogram(binwidth = 1)

summary(wqr$free.sulfur.dioxide)
```

From the histogram above it can be observed that distribution for free sulfur is right skewed wiht few outliers


```{r echo=FALSE}
ggplot(data = wqr, aes(x = total.sulfur.dioxide)) +
geom_histogram(binwidth = 2.5)

summary(wqr$total.sulfur.dioxide)
```

From the histogram above it can be observed that distribution for density is right skewed wiht few outliers


```{r echo=FALSE}
ggplot(data = wqr, aes(x = density)) +
geom_histogram(binwidth = 0.00025)

summary(wqr$density)
```

From the histogram above it can be observed that density is normally distributed


```{r echo=FALSE}
ggplot(data = wqr, aes(x = pH)) +
geom_histogram()

summary(wqr$pH)
```

From the histogram above it can be observed that pH is normally distributed

```{r echo=FALSE}
ggplot(data = wqr, aes(x = sulphates)) +
geom_histogram(binwidth = 0.1)

summary(wqr$sulphates)
```

From the histogram above it can be observed that distribution for sulphates is right-skewed.

```{r echo=FALSE}
ggplot(data = wqr, aes(x = alcohol)) +
geom_histogram()

summary(wqr$alcohol)
```

From the histogram above it can be observed that distribution for alcohol is right-skewed.


Some of the plots are redone wiht better settings
```{r echo=FALSE}
wqr <- wqr[wqr$fixed.acidity < quantile(wqr$fixed.acidity, 0.99),]
wqr <- wqr[wqr$volatile.acidity < quantile(wqr$volatile.acidity, 0.99),]
wqr <- wqr[wqr$residual.sugar < quantile(wqr$residual.sugar, 0.95),]
wqr <- wqr[wqr$chlorides < quantile(wqr$chlorides, 0.95),]
wqr <- wqr[wqr$free.sulfur.dioxide < quantile(wqr$free.sulfur.dioxide, 0.99),]
wqr <- wqr[wqr$total.sulfur.dioxide < quantile(wqr$total.sulfur.dioxide, 0.99),]
wqr <- wqr[wqr$sulphates < quantile(wqr$sulphates, 0.99),]

summary(wqr)

str(wqr)
```


```{r echo=FALSE}
ggplot(data = wqr, aes(x = fixed.acidity)) +
geom_histogram(binwidth = 0.01) +
scale_x_log10()

ggplot(data = wqr, aes(x = volatile.acidity)) +
geom_histogram(binwidth = 0.015) +
scale_x_log10()

ggplot(data = wqr, aes(x = residual.sugar)) +
geom_histogram(binwidth = 0.1)

ggplot(data = wqr, aes(x = chlorides)) +
geom_histogram(binwidth = 0.005)

ggplot(data = wqr, aes(x = free.sulfur.dioxide)) +
geom_histogram(binwidth = 1)

ggplot(data = wqr, aes(x = total.sulfur.dioxide)) +
geom_histogram(binwidth = 2.5)

ggplot(data = wqr, aes(x = sulphates)) +
geom_histogram(binwidth = 0.01) +
scale_x_log10()

```

When plotted on a base 10 logarithmic scale, fixed.acidity, volatile.acidity and sulphates appear to be normally-distributed.
Both residual.sugar and chlorides have left-skewed distrubutions and long tails at the right side if we remove outliers (95%),  both shows normal distributions. Outliners were removed as well for fixed.acidity, volatile.acidity, free.sulfur.dioxide, total.sulfur.dioxide & sulphates.




# Univariate Analysis

### What is the structure of your dataset?
There are 1599 observation and 13 variablas in the dataset. There is one categorical/dependant variable - quality and 11 independant variables.

### What is/are the main feature(s) of interest in your dataset?
The main interest of this analysis is to determine which chemical properties influence the quality of red wines.

###What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
Some of the independent variables could have strong correlation to eacher other but not to quality. This need to be investigated.  

### Did you create any new variables from existing variables in the dataset?
No, I did not.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
The variable X is the series of all observations and will be excluded from furhter analysis.
When plotted on a base 10 logarithmic scale, fixed.acidity, volatile.acidity and sulphates appear to be normally-distributed.
Both residual.sugar and chlorides have left-skewed distrubutions and long tails at the right side if we remove outliers (95%),  both shows normal distributions. Outliners (99%) were removed as well for fixed.acidity, volatile.acidity, free.sulfur.dioxide, total.sulfur.dioxide & sulphates.


# Bivariate Plots Section
First, we will look at correlations between all variablas and based on calculated correlations I will explore strongest correlations among them.

```{r echo=FALSE}
wqr_cor <- wqr[,1:length(wqr)]
round(cor(wqr_cor),3)
```

The independent variables show none to moderate correlations with the dependent variable quality. The variable having the highest correlation of 0.512 with quality is alcohol.


```{r echo=FALSE}
ggpairs(wqr, 
lower = list(continuous = wrap("smooth", shape = I('.'))),
upper = list(combo = wrap("box", outlier.shape = I('.'))))
 
```

Since the objective of this analysis is to explore which chemical components have influence in red wines quality, three of the strongest positive and one negative correlations are explired and ploted.

```{r echo=FALSE}
p1 = ggplot(data = wqr, aes(x = quality, y = alcohol)) +
geom_jitter(alpha = 0.2, position = position_jitter(height = .25)) +
geom_smooth( method = 'lm')

p2 = ggplot(data = wqr, aes(x = quality, y = sulphates)) +
geom_jitter(alpha = 0.2, position = position_jitter(height = .25)) +
geom_smooth( method = 'lm')

p3 = ggplot(data = wqr, aes(x = quality , y = volatile.acidity)) +
geom_jitter(alpha = 0.2, position = position_jitter(height = .25)) +
geom_smooth( method = 'lm')

p4 = ggplot(data = wqr, aes(x = quality , y = citric.acid)) +
geom_jitter(alpha = 0.2, position = position_jitter(height = .25)) +
geom_smooth( method = 'lm')

grid.arrange(p1, p2, p3, p4)
```
  
  alcohol and quality have a strong positive correlation of 0.512
  sulphates and quality have a strong positive correlation of 0.410
  citric.acid and quality have a low positive correlation of 0.228  
  volatile.acidity and quality have a mild negative correlation of -0.367


```{r echo=FALSE}
p1 = ggplot(data = wqr, aes(x = fixed.acidity, y = pH)) +
geom_point(alpha = 0.2) +
geom_smooth( method = 'lm')+
scale_x_log10()

p2 = ggplot(data = wqr, aes(x = citric.acid, y = fixed.acidity)) +
geom_point(alpha = 0.2 ) +
geom_smooth( method = 'lm')

p3 = ggplot(data = wqr, aes(x = free.sulfur.dioxide,
y = total.sulfur.dioxide)) +
geom_point(alpha = 0.2) +
geom_smooth( method = 'lm')

p4 = ggplot(data = wqr, aes(x = fixed.acidity, y = density)) +
geom_point(alpha = 0.2) +
geom_smooth( method = 'lm')

grid.arrange(p1,p2,p3,p4)
```

pH and fixed.acidity have a strong negtive correlation of -0.703
fixed.acidity and citric.acid have a strong positive correlation of 0.689
total.sulfur.dioxide and free.sulfur.dioxide have a strong positive correlation of 0.656
density and fixed.acidity also have a strong positive correlation of 0.661



```{r}
ggplot(data=wqr, aes(x=quality,y=alcohol)) +
geom_jitter(alpha = 0.2, color = "black",
            position = position_jitter(width = .2)) +
geom_line(aes(colour = "Mean"), stat = "summary", fun.y = mean) +
geom_line(aes(colour = "Median"), stat = "summary", fun.y = median) +
  scale_colour_manual("Line colors", values = c("green", "blue"))

```

The mean/median plot of alcohol vs quality shows linear correlation.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Correlation between dependant and independent variables:
  alcohol and quality have a strong positive correlation of 0.512
  sulphates and quality have a strong positive correlation of 0.410
  citric.acid and quality have a low positive correlation of 0.228  
  volatile.acidity and quality have a mild negative correlation of -0.367 

Correlations between independent variables:
  pH and fixed.acidity have a strong negative correlation of -0.703
  fixed.acidity and citric.acid have a strong positive correlation of 0.689
  total.sulfur.dioxide and free.sulfur.dioxide have a strong positive correlation of 0.656
  density and fixed.acidity also have a strong positive correlation of 0.661

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
There are strongcorrelations between pH and fixed.acidity, fixed.acidity and citric.acid, total.sulfur.dioxide and free.sulfur.dioxide & density and fixed.acidity

### What was the strongest relationship you found?
alcohol and quality have the strongest correlation between independent and dependent variables, while pH and fixed.acidity have the strongest interior correlation among independent variables.


# Multivariate Plots Section

We will analyse all the combinations of the most promising variables facet wiht quality. Variables: alcohol,sulphates, citric.acid & volatile.acidity. Those varibles have best correlation factor when compared against quality.

```{r echo=FALSE}
ggplot(data = wqr, aes(x = log10(sulphates), y = alcohol)) +
geom_point(alpha = 0.2) +
facet_wrap(~quality) +
geom_smooth( method = 'lm')
```

Sulphates & alcohol variables facet wiht quality


```{r echo=FALSE}
ggplot(data = wqr, aes(x = volatile.acidity, y = alcohol)) +
geom_point(alpha = 0.2) +
facet_wrap(~quality) +
geom_smooth( method = 'lm')
```

volatile.acidity & alcohol variables facet wiht quality


```{r echo=FALSE}
ggplot(data = wqr, aes(x = citric.acid, y = alcohol)) +
geom_point(alpha = 0.2) +
facet_wrap(~quality) +
geom_smooth( method = 'lm')
```

citric.acid & alcohol variables facet wiht quality


```{r echo=FALSE}
ggplot(data = wqr, aes(x = volatile.acidity, y = log10(sulphates))) +
geom_point(alpha = 0.2) +
facet_wrap(~quality) +
geom_smooth( method = 'lm')
```

volatile.acidity & sulphates variables facet wiht quality


```{r echo=FALSE}
ggplot(data = wqr, aes(x = citric.acid, y = log10(sulphates))) +
geom_point(alpha = 0.2) +
facet_wrap(~quality) +
geom_smooth( method = 'lm')
```

citric.acid & sulphates variables facet wiht quality


```{r echo=FALSE}
ggplot(data = wqr, aes(x = citric.acid, y = volatile.acidity)) +
geom_point(alpha = 0.2) +
facet_wrap(~quality) +
geom_smooth( method = 'lm')

```

citric.acid & volatile.acidity variables facet wiht quality


##Linear Model
Linear model created from four major variables: alcohol, sulphates, citric acid, and volatile acidity. These were all compared to quality.

```{r echo=FALSE}

m1 <- lm(as.numeric(quality) ~ alcohol, data = wqr)
m2 <- update(m1, ~ . + sulphates)
m3 <- update(m2, ~ . + volatile.acidity)
m4 <- update(m3, ~ . + citric.acid)
mtable(m1,m2,m3,m4)
```

Linear model using four of the variables: alcohol, sulphates, volatile acidity & citric acid. R-squared for first three variables is 0.378 and was not changed with forth variable. 


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
Within all independent variables alcohol shows biggest correlation but still weak in correlations with the dependent variable of quality. Independent variables sulphates, volatile.acidity and citric.acid strengthen the correlations. Having a low sulphate count gave the alcohol lower quality. Lower quality wines tended to be lower in alcohol content and citric acid. 

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I created a linear model using four of the variables: alcohol, sulphates, volatile acidity & citric acid. R-squared for first three variables is 0.378 and was not changed with forth variable. I can conclude that linear model is not appropriate method since the R value is too low to be considered valid.
------


# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data = wqr, aes(x = quality, y = alcohol)) +
geom_jitter(alpha = 0.2, position = position_jitter(height = .25), color = 'blue') +
geom_smooth( method = 'lm', color = 'red') +
ggtitle('Alcohol vs Wine Quality') +
xlab('Quality') +
ylab('Alcohol (% by volume)')
```

alcohol and quality have a strong positive correlation of 0.512.
It can be seen from the plot that there is a linear correlation between those two variable. 


### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(data=wqr, aes(x=quality,y=alcohol)) +
geom_jitter(alpha = 0.2, color = "black",
            position = position_jitter(width = .2)) +
geom_line(aes(colour = "Mean"), stat = "summary", fun.y = mean) +
geom_line(aes(colour = "Median"), stat = "summary", fun.y = median) +
  scale_colour_manual("Line colors", values = c("green", "blue")) + 
  ggtitle('Alcohol vs Wine Quality') +
xlab('Quality') +
ylab('Alcohol (% by volume)') + scale_fill_continuous(guide = "legend")
```

It can be seen from the plot that there is a linear correlation for mean and median when observing alcohol and qulity as variables.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = wqr, aes(x = log10(sulphates), y = alcohol)) +
geom_point(alpha = 0.2, color = 'red') +
facet_wrap(~quality) +
geom_smooth( method = 'lm') +
xlab('Sulphates (potassium sulphate - g / dm3)') +
ylab('Alcohol (% by volume)') +
ggtitle('Alcohol vs Sulphates faced by quality')
```

Sulphates & alcohol variables facet wiht quality. Within all independent variables alcohol shows biggest correlation but still weak in correlations with the dependent variable of quality. Independent variables sulphates strengthen the correlation.

------

# Reflection
In this red wine data analysis, I have explored main chemicals which determine the wine quality. It is obvious that chemicals are not the only factors that determine the wine scores but in this analysis we will limit ourselves only to exploring the chemicals.

First part of the analysis was focused on exploration of distributions of each chemical. Here are some observations:
When plotted on a base 10 logarithmic scale, fixed.acidity, volatile.acidity and sulphates appear to be normally-distributed.
Both residual.sugar and chlorides have left-skewed distrubutions and long tails at the right side if we remove outliers (95%),  both shows normal distributions. Outliners (99%) were removed as well for fixed.acidity, volatile.acidity, free.sulfur.dioxide, total.sulfur.dioxide

Second part of analysis was about correlations between all variables. Furthermore the strongest correlations among them were explored.

Correlation between dependant and independent variables:
alcohol and quality have a strong positive correlation of 0.512
sulphates and quality have a strong positive correlation of 0.410
citric.acid and quality have a low positive correlation of 0.228  
volatile.acidity and quality have a mild negative correlation of -0.367 
Correlations between independent variables:
pH and fixed.acidity have a strong negative correlation of -0.703
fixed.acidity and citric.acid have a strong positive correlation of 0.689
total.sulfur.dioxide and free.sulfur.dioxide have a strong positive correlation of 0.656
density and fixed.acidity also have a strong positive correlation of 0.661

Main chemical which have strongest correlation to quality is alcohol content. Alcohol variable itself is not sufficient but rather in combination with sulphates and acidity. The correlations for these variables are within reasonable bounds. 

In multivariate analysis the linear model was created using four of the variables: alcohol, sulphates, volatile acidity & citric acid. R-squared for first three variables is 0.378 and was not changed with forth variable. Conclusion was that created linear model is not appropriate method since the R value is too low to be considered valid.

Furthermore, the weaknesses of this data is due to biases of the sommelier way of determining the quality not based on the chemicals. I assume that sommeliers always look for the same features and therefore the grades could be similar. Also I would assume that different sommeliers schools could possibly teach different techniques of wine grading. It would be nice to compare grades on same wines by sommeliers from different continents.
Overall conclusion is that chemicals are not the only factor that have to be looked at when determine the quality at the wine.


